A generalization of the Bradley-Terry model for draws in chess

### 2019-05-27 inst/collusion_chess_draws.Rnw
more work, give ghosts a fraktur font

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}

@@

\subsection{First-mover advantage} Traditionally White moves first in
any game, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of~\cite{adams2007}---``white
to play and win''---to~\citet{adorjan1988}---``black is OK!''.

@@

Here I show how Bradley-Terry may be generalized ot account for the
existence of draws and the first-mover advantage.  Consider first the
standard Bradley-Terry model for a finite number of chess players.
Ignoring the effect of playing white, we wish to find numbers
(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
assumption of independence of trials, would be

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and a third entity (a
``draw monster'') that wins if the game is drawn.  The system is
further modified to account for first-move advantage by positing an
entity whose strength is added to that of the white player.  This
methodology affords insight into players' strengths, response to
playing black or white, and risk-aversion as manifested by probability
of drawing.  If we consider a chess game to be a three-way trial, an
appropriate likelihood function would be for non-negative
$p_1,\ldots,p_n,D$ ($D$ being the draw monster's strength), with
$D+\sum p_i=1$ and

\begin{equation}
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j+D}\right)^{W_{ij}}
\left(\frac{p_j}{p_i+p_j+D}\right)^{L_{ij}}
\left(\frac{D  }{p_i+p_j+D}\right)^{D_{ij}}
\end{equation}

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength $W$ is
added to the white player.  The strength of the white player is thus
the sum of the player's strength and $W$.  Recalling that a game might
be a draw, we obtain the probabilities as given in
table~\ref{probabilities}.  The corresponding likelihood function is
given below in equation \ref{likelihood_WD}:

@@

\begin{table}
\centering
\begin{tabular}{l|cc}\\
  &$i$ plays white & $j$ plays white\\ \hline
$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+W\oo}{p_i+p_j+W+D}$ & $\frac{p_i}{p_i+p_j+W+D}$  \\
$\Prob{(\mbox{draw})}$      & $\frac{D\oo}{p_i+p_j+W+D}$     & $\frac{D}{p_i+p_j+W+D}$    \\
$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+W+D}$   & $\frac{p_j+W}{p_i+p_j+W+D}$\\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two players under a generalized
  Bradley-Terry model\label{probabilities}}
\end{table}

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+W}{p_i+p_j+D+W}  \right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+D+W    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+D+W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
\left(\frac{p_j+W}{p_i+p_j+D+W  }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{D}{p_i+p_j+D    }\right)^{D_{ij}}
\nonumber
\end{align}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{black}}
=
\prod_{i<j}
\left(\frac{D}{p_i+p_j+D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

@@

The above likelihood functions are objective and reasonable
representations of chess game outcomes.  However, they differ markedly
from previous likelihood functions of draws in that they include
reified entities representing proclivity to draw $D$, and white or
first-mover advantage ($W$).

@@

\cite{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$D=\nu\sqrt{p_ip_j}$.  However, \cite{joe1990} uses likelihood methods
to show that this model does not fit data well.  Here we note that
such a model automatically give stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.  \cite{moul2009} consider chess draws and use sophisticated
regression techniques to analyse games in terms of not only the
result, but also the ply-count, data not used in the present analysis.

@@

\begin{equation}\label{kka_likelihood}
\mathcal{L}\left(p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},W,D\right)=
\frac{
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12} 
p_\mathrm{Ks}^{20}
D^{224}
\left(p_\mathrm{Kr} + W\right)^{25}
\left(p_\mathrm{Ks} + W\right)^{45}
\left(p_\mathrm{An} + W\right)^{24}
}{
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + D\right)^{193}
\left(p_\mathrm{Kr} + p_\mathrm{An} + W + D\right)^{92}
\left(p_\mathrm{Ks} + p_\mathrm{An} + W + D\right)^{80}
}
\end{equation}

@@

This likelihood function is easily maximized numerically and the
evaluate is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of
equation~\ref{kka_likelihood}.  From the figure, we see that the
maximum likelihood estimate for draw strength $\hat{D}\simeq 0.53$:
compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 games.  The na\"{i}ve estimate is lower
because the maximization process automatically adjusts for the
different players' strengths.  Secondly, we may make inferences about
$W$, the extra strength conferred by playing white: it is conceivable
that $W=0$ as this term does not appear on its own in the numerator of
any term.  We may test $H_0\colon W=0$ using the Method of Support by
maximizing the likelihood subject to the constraint $W=0$ and
comparing with the likelihood when the optimization is free.

@@

The result of numerical optimization is about 7.01 units of support,
or a likelihood ratio of $e^{7.01}\simeq 1107.7$: that is, well over
the two-units-of-support per degree of freedom suggested by
\citeauthor{edwards1972}; we can convincingly reject the hypothesis
that $W=0$ and infer that playing white confers a significant
advantage.  This is consistent with expectation, but it is now
possible to go significantly beyond mere existence of the effect and
to quantify it, accounting for the differential strengths of the
players.  Taking the MLE above, and considering a game between say
Karpov and Anand, the strengths are Karpov=0.064, Anand=0.151, white
=0.104 and draw =0.53.  Thus we would estimate that Karpov playing
white against Kasparov would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\ {=}0.627\ {-}0.297$ if
Karpov plays black).

@@

As argued above, players with a cautious style tend to draw and this
is an important and interesting component of the game.  To investigate
this we need a likelihood function that can account for possible
differences in draw proclivity among the players.  This may be
achieved by considering a game between two players to be a contest
between five entities: the two real players, the white advantage, and
two {\em personalised} draw entities, one for each real player; a
drawn game is counted as a win for either of the draw ``ghosts''.  Under
these assumptions, table~\ref{kka} translates into the following
likelihood function:

@@

\begin{equation}\label{KKA_personalised_drawmonster}
\mathcal{L}\left(
p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
,W,
d_\mathrm{An},d_\mathrm{Kr},d_\mathrm{Ks}
\right)=
\frac{
\parbox{3in}{$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
\left(p_\mathrm{Kr} + W\right)^{25}
\left(p_\mathrm{Ks} + W\right)^{45} 
\left(p_\mathrm{An} + W\right)^{24}\\ \rule{10mm}{0mm}
\left(d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{129}
\left(d_\mathrm{Kr} + d_\mathrm{An}\right)^{49}
\left(d_\mathrm{Ks} + d_\mathrm{An}\right)^{46}
$}}{\parbox{3in}{$
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{An} + W + d_\mathrm{Kr} + d_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Ks} + p_\mathrm{An} + W + d_\mathrm{Ks} + d_\mathrm{An}\right)^{80}
$}}
\end{equation}

@@

In figure~\ref{kka_3draws}, we see considerable variability in
proclivity to draw.  Again this may be tested with the Method of
Support, which gives a $p$-value of about 0.039, borderline
significant: an interesting result, given that in this case no draw
monster appears alone anywhere in the likelihood function.

@@

\subsection{Each player has a distinct White Monster}

We now consider the possibility that the strength conferred by playing
white, is specific to each player and we denote this $w_\mathrm{An}$,
$w_\mathrm{Kr}$ and $w_\mathrm{Ks}$.  The corresponding likelihood
function would be

\begin{equation}\label{KKA_personalised_whitemonster}
%\mathcal{L}\left(
%p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},
%w_\mathrm{An},w_\mathrm{Kr},w_\mathrm{Ks},D
%\right)=
\frac{
%\parbox{3in}
%       {$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
D^{224}
\left(p_\mathrm{An} + w_\mathrm{An}\right)^{24}%\\ \rule{10mm}{1mm}
\left(p_\mathrm{Kr} + w_\mathrm{Kr}\right)^{25}
\left(p_\mathrm{Ks} + w_\mathrm{Ks}\right)^{45}
%$}
}{\parbox{4in}{$
\left(p_\mathrm{An} + p_\mathrm{Kr} + w_\mathrm{An} +  D\right)^{43}
\left(p_\mathrm{An} + p_\mathrm{Kr} + w_\mathrm{Kr} +  D\right)^{49}\\ \rule{10mm}{0mm}
\left(p_\mathrm{An} + p_\mathrm{Ks} + w_\mathrm{An} +  D\right)^{37}
\left(p_\mathrm{An} + p_\mathrm{Ks} + w_\mathrm{Ks} +  D\right)^{43}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + w_\mathrm{Kr} +  D\right)^{99}
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + w_\mathrm{Ks} +  D\right)^{94}
$}}
\end{equation}

In the likelihood function above we see, for example, that Anand won a
total of $18+6=24$ games playing white, $13+2=15$ games playing
Black, and that Anand played white against Karpov a total of
$18+20+5=43$ times.  We fail to reject the null of
$w_\mathrm{An}=w_\mathrm{Kr}=w_\mathrm{Ks}$, having a likelihood ratio
of $4.5$, or an asymptotic $p$-value of $0.076$.  There is no strong
evidence that the players differ in regard to their white advantage.

@@

Here I consider game results from Cura\c{c}ao 1962, reproduced in
Table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher's exact test, one-sided,
$p=0.0079$).  To quantify such Soviet collusion, accounting for the
player skills, we introduce a new draw entity: a collusive draw
monster $C$, who wins when two colluding players draw, and is inactive
whenever a game has at least one non-colluding player.  Then
$H_N\colon C=D$ corresponds to no collusion and the likelihood
function would have

\begin{equation}\parbox{4in}{$
P^5\,
{Ke}^5\,
G^4
\mathit{Fis}\,  % Fischer, cf Filip
\mathit{Ko}^5
B
T^2  % Tal
\mathit{Fil}\,
D^{42}C^{12}\oo\\ \rule{10mm}{0mm}
\left(P+W\right)^3
\left(\mathit{Ke}+W\right)^4
\left(G+W\right)^4\
\left(\mathit{Fis}+W\right)^7\oo\\ \rule{20mm}{0mm}
\left(\mathit{Ko}+W\right)^2
\left(B+W\right)^5
\left(T+W\right)
\left(\mathit{Fil}+W\right)\oo
$}\end{equation}

\noindent for the numerator (where $C$ is the strength of a collusive
draw), and a total of 30 terms such as
$\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
  Keres 4 times] in the denominator.  The likelihood ratio for the
hypothesis that collusive draws have the same strength as regular
draws is about $e^{11.8}\simeq 10^5$: we conclude that the high number
of draws among the three accused players was unlikely to be due to
chance.

@@

\subsection{Chess collusion: interzonal Stockholm 1962}

@@

The fifth Interzonal took place in Stockholm in 1962, in the form of a
23-player round-robin tournament.  For the results shown in
table~\ref{interzonal}, we would have a likelihood function

\begin{equation}
\mathcal{L}\left(F,G,P,K,L,W,D\right)=
%                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
%                             D/(G+P+D+W)         D/(G+K+W+D)       D/(G+F+W+D)
%                                                 D/(P+K+W+D)       D/(P+L+W+D)
%                                                                   (K+W)/(K+L+W+D)
\frac{
  D^8(F+W)(K+W)
}{\parbox{4in}{$
    (F+G+W+D)(F+P+W+D)\\ \rule{10mm}{0mm}(F+K+W+D)  (F+L+W+D)\\ \rule{20mm}{0mm}
         (G+P+W+D)  (G+K+W+D)\\ \rule{30mm}{0mm}(G+F+W+D)
                    (P+K+W+D)\\ \rule{40mm}{0mm}(P+L+W+D)
                               (K+L+W+D)
 $}}
\end{equation}

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of increased Soviet drawing proclivity, we
introduce a Soviet draw monster with strength~$D_s$, who wins when two
Soviet players draw, and is inactive if a game has at least one
non-Soviet player.  Then $H_N\colon D_s=D$ corresponds to no collusion
and the likelihood function for table~\ref{interzonal} would be

\begin{equation}\label{collusion}
\mathcal{L}\left(F,G,P,K,L,W,D,D_s\right)=
%                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
%                             D/(G+P+D+W)         D/(G+K+W+D)       D/(G+F+W+D)
%                                                 D/(P+K+W+D)       D/(P+L+W+D)
%                                                                   (K+W)/(K+L+W+D)
\frac{
  D_s^3D^5(F+W)(K+W)
}{\parbox{4in}{$
    (F+G+W+D)(F+P+W+D)\\ \rule{10mm}{0mm}(F+K+W+D)  (F+L+W+D)\\ \rule{20mm}{0mm}
         (G+P+W+D_s)  (G+K+W+D_s)\\ \rule{30mm}{0mm}(G+F+W+D)
                    (P+K+W+D_s)\\ \rule{40mm}{0mm}(P+L+W+D)
                               (K+L+W+D)
 $}}
\end{equation}

The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a
constrained maximization gives a likelihood ratio of about 5096, or a
$p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further
and plot a profile likelihood curve for the log-odds ratio
$\log\left(D_s/D\right)$ which, if Dirichlet, would be accurately
parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
illustrated in figure~\ref{proflike}, that shows a credible interval
of about 1.8-2.4.

>>>>

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
\newcommand{\white}{\ensuremath{\mathfrak{W}}}

@@

\subsection{First-mover advantage} Traditionally White moves first in
any game, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of~\cite{adams2007}---``White
to play and win'' to~\citet{adorjan1988}---``Black is OK!''.

@@

Here I show how Bradley-Terry may be generalized to account for the
existence of draws and the first-mover advantage.  Consider first the
standard Bradley-Terry model for a finite number of chess players.
Ignoring the effect of playing white, we wish to find numbers
(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
assumption of independence of trials, would be

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and a third entity (a
``draw monster'') that wins if the game is drawn.  The system is
further modified to account for first-move advantage by positing an
entity whose strength is added to that of the white player.  This
methodology affords insight into players' strengths, response to
playing black or white, and risk-aversion as manifested by probability
of drawing.  If we consider a chess game to be a three-way trial, an
appropriate likelihood function would be for non-negative
$p_1,\ldots,p_n,\draw$ (here \draw\ is the draw monster's strength,
fraktur font being used for reified entities' strengths) with $\draw+\sum
p_i=1$ and

\begin{equation}
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j+\draw}\right)^{W_{ij}}
\left(\frac{p_j}{p_i+p_j+\draw}\right)^{L_{ij}}
\left(\frac{\draw  }{p_i+p_j+\draw}\right)^{D_{ij}}
\end{equation}

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength \white\ is
added to the white player.  The strength of the white player is thus
the sum of the player's strength and \white.  Recalling that a game
might be a draw, we obtain the probabilities as given in
Table~\ref{probabilities}.  The corresponding likelihood function is
given below in equation \ref{likelihood_WD}:

@@

\begin{table}
\centering
\begin{tabular}{l|cc}\\
  &$i$ plays white & $j$ plays white\\ \hline
$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+\white\oo}{p_i+p_j+\white+\draw}$ & $\frac{p_i}{p_i+p_j+\white+\draw}$  \\
$\Prob{(\mbox{draw})}$      & $\frac{\draw\oo}{p_i+p_j+\white+\draw}$     & $\frac{\draw}{p_i+p_j+\white+\draw}$    \\
$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+\white+\draw}$   & $\frac{p_j+\white}{p_i+p_j+\white+\draw}$\\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two players under a generalized
  Bradley-Terry model\label{probabilities}}
\end{table}

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+\white}{p_i+p_j+\draw+\white}  \right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+\draw+\white    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+\draw+\white}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
\left(\frac{p_j+\white}{p_i+p_j+\draw+\white  }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{\draw}{p_i+p_j+\draw    }\right)^{D_{ij}}
\nonumber
\end{align}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{\draw}{p_i+p_j+\draw+\white}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{\draw}{p_i+p_j+\draw+\white}\right)^{D_{ij}^\mathrm{black}}
=
\prod_{i<j}
\left(\frac{\draw}{p_i+p_j+\draw}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

@@

The above likelihood functions are objective and reasonable
representations of chess game outcomes.  However, they differ markedly
from previous likelihood functions of draws in that they include
reified entities representing proclivity to draw \draw, and white or
first-mover advantage \white.

@@

\cite{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$\draw=\nu\sqrt{p_ip_j}$.  However, \cite{joe1990} uses likelihood methods
to show that this model does not fit data well.  Here we note that
such a model automatically give stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.  \cite{moul2009} consider chess draws and use sophisticated
regression techniques to analyse games in terms of not only the
result, but also the ply-count, data not used in the present analysis.

@@

\begin{equation}\label{kka_likelihood}
%\mathcal{L}\left(p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},\white,\draw\right)=
\frac{
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12} 
p_\mathrm{Ks}^{20}
\draw^{224}
\left(p_\mathrm{Kr} + \white\right)^{25}
\left(p_\mathrm{Ks} + \white\right)^{45}
\left(p_\mathrm{An} + \white\right)^{24}
}{
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white + \draw\right)^{193}
\left(p_\mathrm{Kr} + p_\mathrm{An} + \white + \draw\right)^{92}
\left(p_\mathrm{Ks} + p_\mathrm{An} + \white + \draw\right)^{80}
}.
\end{equation}

@@

This likelihood function is easily maximized numerically and the
evaluate is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $\draw=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of
equation~\ref{kka_likelihood}.  From the figure, we see that the
maximum likelihood estimate for draw strength $\hat{\draw}\simeq 0.53$:
compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 games.  The na\"{i}ve estimate is lower
because the maximization process automatically adjusts for the
different players' strengths.  Secondly, we may make inferences about
$\white$, the extra strength conferred by playing white: it is conceivable
that $\white=0$ as this term does not appear on its own in the numerator of
any term.  We may test $H_0\colon \white=0$ using the Method of Support by
maximizing the likelihood subject to the constraint $\white=0$ and
comparing with the likelihood when the optimization is free.

@@

The result of numerical optimization is about 7.01 units of support,
or a likelihood ratio of $e^{7.01}\simeq 1107.7$: that is, well over
the two-units-of-support per degree of freedom suggested by
\citeauthor{edwards1972}; we can convincingly reject the hypothesis
that $\white=0$ and infer that playing white confers a significant
advantage.  This is consistent with expectation, but it is now
possible to go significantly beyond mere existence of the effect and
to quantify it, accounting for the differential strengths of the
players.  Taking the MLE above, and considering a game between say
Karpov and Anand, the strengths are Karpov=0.064, Anand=0.151, \white
=0.104 and \draw =0.532.  Thus we would estimate that Karpov playing
white against Kasparov would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\ {=}0.627\ {-}0.297$ if
Karpov plays black).

@@

As argued above, players with a cautious style tend to draw and this
is an important and interesting component of the game.  To investigate
this we need a likelihood function that can account for possible
differences in draw proclivity among the players.  This may be
achieved by considering a game between two players to be a contest
between five entities: the two real players, the white advantage, and
two {\em personalised} draw entities, one for each real player; a
drawn game is counted as a win for either of the draw monsters.  Under
these assumptions, table~\ref{kka} translates into the following
likelihood function:

@@

\begin{equation}\label{KKA_personalised_drawmonster}
%\mathcal{L}\left(
%p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
%,\white,
%\draw_\mathrm{An},\draw_\mathrm{Kr},\draw_\mathrm{Ks}
%\right)=
\frac{
\parbox{3in}{$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
\left(p_\mathrm{Kr} + \white\right)^{25}
\left(p_\mathrm{Ks} + \white\right)^{45} 
\left(p_\mathrm{An} + \white\right)^{24}\\ \rule{10mm}{0mm}
\left(\draw_\mathrm{Kr} + \draw_\mathrm{Ks}\right)^{129}
\left(\draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{49}
\left(\draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{46}
$}}{\parbox{3in}{$
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{An} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Ks} + p_\mathrm{An} + \white + \draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{80}
$}}
\end{equation}

@@

In Figure~\ref{kka_3draws}, we see considerable variability in
proclivity to draw.  Again this may be tested with the Method of
Support, which gives a $p$-value of about 0.039, borderline
significant: an interesting result, given that in this case no draw
monster appears alone anywhere in the likelihood function.

@@

\subsection{Each player has a distinct white monster}

We now consider the possibility that the strength conferred by playing
white is specific to each player and we denote the strengths
$\white_\mathrm{An}$, $\white_\mathrm{Kr}$ and $\white_\mathrm{Ks}$.
The corresponding likelihood function would be

\begin{equation}\label{KKA_personalised_whitemonster}
%\mathcal{L}\left(
%p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},
%\white_\mathrm{An},\white_\mathrm{Kr},\white_\mathrm{Ks},\draw
%\right)=
\frac{
%\parbox{3in}
%       {$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
\draw^{224}
\left(p_\mathrm{An} + \white_\mathrm{An}\right)^{24}%\\ \rule{10mm}{1mm}
\left(p_\mathrm{Kr} + \white_\mathrm{Kr}\right)^{25}
\left(p_\mathrm{Ks} + \white_\mathrm{Ks}\right)^{45}
%$}
}{\parbox{4in}{$
\left(p_\mathrm{An} + p_\mathrm{Kr} + \white_\mathrm{An} +  \draw\right)^{43}
\left(p_\mathrm{An} + p_\mathrm{Kr} + \white_\mathrm{Kr} +  \draw\right)^{49}\\ \rule{10mm}{0mm}
\left(p_\mathrm{An} + p_\mathrm{Ks} + \white_\mathrm{An} +  \draw\right)^{37}
\left(p_\mathrm{An} + p_\mathrm{Ks} + \white_\mathrm{Ks} +  \draw\right)^{43}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white_\mathrm{Kr} +  \draw\right)^{99}
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white_\mathrm{Ks} +  \draw\right)^{94}
$}}
\end{equation}

In the likelihood function above we see, for example, that Anand won a
total of $18+6=24$ games playing white, $13+2=15$ games playing Black,
and that Anand played white against Karpov a total of $18+20+5=43$
times.  We fail to reject the null of
$\white_\mathrm{An}=\white_\mathrm{Kr}=\white_\mathrm{Ks}$, having a
likelihood ratio of $4.5$, or an asymptotic $p$-value of $0.076$.
There is no strong evidence that the players differ in regard to their
white advantage.

@@

Here I consider game results from Cura\c{c}ao 1962, reproduced in
Table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher's exact test, one-sided,
$p=0.0079$).  To quantify such Soviet collusion, accounting for the
player skills, we introduce a new draw entity: a collusive draw
monster with strength \cdraw, who wins when two colluding players
draw, and is inactive whenever a game has at least one non-colluding
player.  Then $H_N\colon\cdraw=\draw$ corresponds to no collusion and
the likelihood function would have

\begin{equation}\parbox{4in}{$
P^5\,
{Ke}^5\,
G^4
\mathit{Fis}\,  % Fischer, cf Filip
\mathit{Ko}^5
B\,
T^2  % Tal
\mathit{Fil}\,
\draw^{42}\,\cdraw^{12}\oo\\ \rule{10mm}{0mm}
\left(P+\white\right)^3
\left(\mathit{Ke}+\white\right)^4
\left(G+\white\right)^4\
\left(\mathit{Fis}+\white\right)^7\oo\\ \rule{20mm}{0mm}
\left(\mathit{Ko}+\white\right)^2
\left(B+\white\right)^5
\left(T+\white\right)
\left(\mathit{Fil}+\white\right)\oo
$}\end{equation}

\noindent for the numerator (where \cdraw\ is the strength of a collusive
draw), and a total of 30 terms such as
$\left(P+\mathit{Ke}+\white+\cdraw\right)^{4}$ [indicating that Petrosian played
  Keres 4 times] in the denominator.  The likelihood ratio for the
hypothesis that collusive draws have the same strength as regular
draws is about $e^{11.8}\simeq 10^5$: we conclude that the high number
of draws among the three accused players was unlikely to be due to
chance.

@@

\subsection{Soviet playing style in the interzonal Stockholm 1962}

@@

The fifth Interzonal took place in Stockholm in 1962, in the form of a
23-player round-robin tournament.  For the results shown in
Table~\ref{interzonal}, we would have a likelihood function

\begin{equation}
%\mathcal{L}\left(F,G,P,K,L,\white,\draw\right)=
%                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
%                             D/(G+P+D+W)         D/(G+K+W+D)       D/(G+F+W+D)
%                                                 D/(P+K+W+D)       D/(P+L+W+D)
%                                                                   (K+W)/(K+L+W+D)
\frac{
  \draw^8(F+\white)(K+\white)
}{\parbox{4in}{$
    (F+G+\white+\draw)(F+P+\white+\draw)\\ \rule{10mm}{0mm}(F+K+\white+\draw)  (F+L+\white+\draw)\\ \rule{20mm}{0mm}
         (G+P+\white+\draw)  (G+K+\white+\draw)\\ \rule{30mm}{0mm}(G+F+\white+\draw)
                    (P+K+\white+\draw)\\ \rule{40mm}{0mm}(P+L+\white+\draw)
                               (K+L+\white+\draw)
 $}}
\end{equation}

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $\draw^{110}$.  To
investigate the possibility of increased Soviet drawing proclivity, we
introduce a Soviet draw monster with strength~$\draw_s$, who wins when two
Soviet players draw, and is inactive if a game has at least one
non-Soviet player.  Then $H_N\colon \draw_s=\draw$ corresponds to no collusion
and the likelihood function for Table~\ref{interzonal} would be

\begin{equation}\label{collusion}
%\mathcal{L}\left(F,G,P,K,L,\white,\draw,\draw_s\right)=
%                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
%                             D/(G+P+D+W)         D/(G+K+W+D)       D/(G+F+W+D)
%                                                 D/(P+K+W+D)       D/(P+L+W+D)
%                                                                   (K+W)/(K+L+W+D)
\frac{
  \draw_s^3\draw^5(F+\white)(K+\white)
}{\parbox{4in}{$    (F+G+\white+\draw)(F+P+\white+\draw)\\ \rule{10mm}{0mm}(F+K+\white+\draw)  (F+L+\white+\draw)\\ \rule{20mm}{0mm}
         (G+P+\white+\draw_s)  (G+K+\white+\draw_s)\\ \rule{30mm}{0mm}(G+F+\white+\draw)
                    (P+K+\white+\draw_s)\\ \rule{40mm}{0mm}(P+L+\white+\draw)
                               (K+L+\white+\draw)
 $}}
\end{equation}
